Type-Safe Modular Parsing

### 2017-09-07 SLE_17/sections/Introduction.tex
Add the reference.

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. Techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining operations
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

@@

\item \textbf{A Parsing Technique for OO ASTs:} A simplified version of
  our technique also enables parsing OO-style ASTs, where new language
  constructs can be easily added.

>>>>

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs. \cite{SLOANE201520}
provides a near solution to modular parsing by using case classes, but
it could be potentially unsafe since exhaustiveness of pattern matching
is not guaranteed. In particular the
problem of how to modularize parsing, including its algorithmic challenges,
has not been studied in other approaches. Techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining operations
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

@@

### 2017-09-07 SLE_17/sections/RelatedWork.tex
Add the reference.

\paragraph{Safely Composable Type-Specific Languages} There is almost no work on 
semantically modular parsing. A notable exception is the work 
on safely composable type-specific languages~\cite{omar14}. In this
work the extensible language Wyvern supports the addition of new
syntax and semantics, while preserving type-safety and separate
compilation. However this approach and our work have different goals: 
their approach is aimed at supporting extensibility of Wyvern with
new syntax; whereas our approach is a general technique aimed at 
modular parsing of any languages. In contrast to their modular parsing 
approach, which is directly built-in to the Wyvern language, our approach is 
library-based and can be used by many  mainstream OO languages.

@@

There are also many design patterns in OO languages that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.
On the other hand we are unware of any work on OOP
that has covered how to do modular parsing for extensible ASTs.

It is worth mentioning that Scala case classes~\cite{emir2007matching} provide a near
solution to the Expression Problem. Case classes can be
modularly added, but they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writing extensible code with case classes. So that full static type-safety is not ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore, modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section~\ref{sec:inheritance} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.
\vspace{-7pt}

>>>>

\paragraph{Modular Parsing with Case Classes}

Our work is close to~\cite{SLOANE201520}, in the sense that they also make use of Packrat parsers
and standard OO inheritance achieving modular parsing. The difference is that case
classes are used in their work for extensibility. Case classes~\cite{emir2007matching} provide a near solution
to the Expression Problem since they can be modular added, and hence are widely used in
practice, but they do not enforce exhaustiveness
of pattern matching for extensible operations. In other words run-time pattern matching errors
can happen when writing extensible code with case classes, so that full static type-safety is not
ensured. Our work uses Object Algebras for full extensibility and type safety, and we have well
studied the algorithmic challenges of parsing in a modular setting.

\paragraph{Safely Composable Type-Specific Languages}

There is almost no work on 
semantically modular parsing. A notable exception is the work 
on safely composable type-specific languages~\cite{omar14}. In this
work the extensible language Wyvern supports the addition of new
syntax and semantics, while preserving type-safety and separate
compilation. However this approach and our work have different goals: 
their approach is aimed at supporting extensibility of Wyvern with
new syntax; whereas our approach is a general technique aimed at 
modular parsing of any languages. In contrast to their modular parsing 
approach, which is directly built-in to the Wyvern language, our approach is 
library-based and can be used by many  mainstream OO languages.

@@

There are also many design patterns in OO languages that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.
\vspace{-7pt}
